---
title: 'Statistical Methods for Discrete Response, Time Series, and Panel Data (W271): Group Lab 3'
geometry: margin=1in
output:
  pdf_document:
    latex_engine: xelatex
  number_sections: yes
  html_document: default
  toc: yes
fontsize: 11pt
---

# U.S. traffic fatalities: 1980-2004

In this lab, you are asked to answer the question **"Do changes in traffic laws affect traffic fatalities?"**  To do so, you will conduct the tasks specified below using the data set *driving.Rdata*, which includes 25 years of data that cover changes in various state drunk driving, seat belt, and speed limit laws. 

Specifically, this data set contains data for the 48 continental U.S. states from 1980 through 2004. Various driving laws are indicated in the data set, such as the alcohol level at which drivers are considered legally intoxicated. There are also indicators for “per se” laws—where licenses can be revoked without a trial—and seat belt laws. A few economics and demographic variables are also included. The description of the each of the variables in the dataset is come with the dataste.

```{r common, warning=FALSE,message=FALSE}
library(foreign) 
library(gplots) 
library(ggplot2) 
library(stats) 
library(Hmisc) 
library(car) 
library(dplyr) 
#library(corrplot) 
#library(corrgram) 
library(lattice) 
library(plm)
```


**Exercises:**

1. (30%) Load the data. Provide a description of the basic structure of the dataset, as we have done throughout the semester. Conduct a very thorough EDA, which should include both graphical and tabular techniques, on the dataset, including both the dependent variable *totfatrte* and the potential explanatory variables. You need to write a detailed narrative of your observations of your EDA. *Reminder: giving an "output dump" (i.e. providing a bunch of graphs and tables without description and hoping your audience will interpret them) will receive a zero in this exercise.*

```{r 1 load, warning=FALSE,message=FALSE}
driving <- miceadds::load.Rdata2( filename="driving.Rdata")

#load data
driving <- get(load('driving.RData'))

#check the variables in the new environment
ls.str(ex)

#describe the variables 
desc

#examine the dat
rbind(head(data,2),tail(data,2))

#check for nulls
apply(data, 2, function(x) any(is.na(x)))

#check for gaps in panel
table(data$state)
table(data$year)

table(data$sl70plus, data$year)

```
The Dataset is panel data, that contains observations about different US states from year 1980 to 2004. There are 1200 observations in total, with 56 columns . The data has 25 observations each, one per year, from 48 states except state ids 2,3 and 9. All variables are observed for all states and over all time periods, hence the panel is balanced.

The fields in the data can be divided into *dimensions*, *attributes* and *measures*. Dimensions are the unique combination that identifies the data record (aka the index of the panel data), which are `Year` and `State`. Attributes are the features of the dimensions, which for us is the traffic laws and demographic variables. Both types - dimensions and attributes - are explanatory variables. Measures are the quantitative variables of interest that includes fatality numbers. For us, the measure is *totfatrte*. 

For EDA, to determine both the common and individual driving behaviors of US States across time, we'll analyze the aggregate of traffic laws in US across time. Then we'll focus on the fatality progression of top and bottom ranked US states across years. We'll also evaluate how the fatality pattern is different between years 1980 and 2004. 

Below we analyze the growth in fatalities over years by state
```{r 1.EDA1,warning=FALSE,message=FALSE}
ggplot(data, mapping = aes(x = year, y = totfatrte)) +
  geom_line(aes(linetype = as.factor(state)))
```



```{r 1 EDA2,warning=FALSE,message=FALSE}
# summarize the average statistics for speed limit in a data frame
sl.df <- data %>% group_by(year) %>%
summarise(sl55 = mean(sl55), sl65 = mean(sl65),sl70 = mean(sl70),sl75 = mean(sl75), slnone = mean(slnone))
sl.df

sl.plot <- ggplot(sl.df, aes(x = year)) + geom_line(aes(x = year, y = sl55, color = "sl55"), size = 1.5, group = 1)  + geom_line(aes(x = year, y = sl65, color = "sl65"), size = 1.5, group = 1) + geom_line(aes(x = year, y = sl70, color = "sl70"), size = 1.5, group = 1) + geom_line(aes(x = year, y = sl75, color = "sl75"), size = 1.5, group = 1) + geom_line(aes(x = year, y = slnone, color = "slnone"), size = 1.5, group = 1) + scale_x_continuous(breaks = seq(min(sl.df$year), max(sl.df$year), 1)) + theme(axis.text.x = element_text(angle = 90, vjust=0.5)) + scale_color_manual(name = "Speed Limit",
      values = c(
           sl55="red",
           sl65="blue",
           sl70="green",
           sl75="yellow",
           slnone="violet" )) + labs(title = "Changes in speed limit law prevalence over time",
y = "[%] states with given speed limit)",
x = "")

sl.plot 

# summarize the average statistics for blood alcohol in a data frame
bac.df <- data %>% group_by(year) %>%
summarise(bac10 = mean(bac10), bac08 = mean(bac08))

bac.plot <- ggplot(bac.df, aes(x = year)) +
geom_line(aes(y = bac10, color='bac10'), size = 1.5, group = 1) + geom_line(aes(y=bac08, color='bac08'), size = 1.5, group = 1) + scale_x_continuous(breaks = seq(min(sl.df$year), max(sl.df$year), 1)) + theme(axis.text.x = element_text(angle = 90, vjust=0.5)) + scale_color_manual(name = "Blood Alcohol Level",
      values = c(
           bac08="red",
           bac10="blue")) + labs(title = "Changes in Blood Alcohol law over time",
y = "[%] states with given Alcohol)",
x = "")

bac.plot 

#summarize general laws 
glaws.df <- data %>% group_by(year) %>%
summarise(zerotol = mean(zerotol), gdl = mean(gdl), perse = mean(perse), seatbelt = mean(seatbelt))

glaws.plot <- ggplot(glaws.df, aes(x = year)) +
geom_line(aes(y = zerotol, color='pink')) + geom_line(aes(y=gdl, color='green')) + geom_line(aes(y=perse, color='blue')) + geom_line(aes(y=seatbelt, color='cyan'))+ scale_x_continuous(breaks = seq(min(sl.df$year), max(sl.df$year), 1)) + theme(axis.text.x = element_text(angle = 90, vjust=0.5)) + scale_color_discrete(name="General Traffic Laws", labels=c("zerotol", "gdl","perse", "seatbelt")) + labs(title = "Changes in General Traffic laws over time",
y = "[%] states with given Laws)",
x = "")

glaws.plot 
```

```{r 1 EDA3}
#fatality change by year 
traffic.yearly.aggr <- data %>%
  group_by(year) %>%
  summarise_at(vars(totfat, nghtfat,wkndfat), funs(sum))

ggplot(traffic.yearly.aggr, aes(year, totfat)) +
  geom_bar(aes(fill = factor(year)), position = "dodge", stat="identity") + ggtitle("Fatalities by year") + theme_gray() + scale_fill_hue(c=45,l=80) + theme(plot.title = element_text(size = 10, hjust = 0.5)) + 
  geom_line(aes(x = year, y = nghtfat, color="red"), size = 1.5, group = 1) + 
geom_line(aes(x = year, y = wkndfat, color="blue"), size = 1.5, group = 1) +
scale_color_discrete(name = "Rates", labels = c("Night Fatality Rate", "Weekend Fatality Rate"))

```

```{r 1 EDA3}
#fatality change by state 

#get state name 
us.states = read.csv("usstates.csv", header = TRUE, sep = ",", dec = ".")

data.with.name <- merge(data, us.states, by=c("state","state"))

traffic.state.aggr <- data.with.name %>%
  group_by(shortname) %>%
  summarise_at(vars(totfat, nghtfat,wkndfat), funs(sum))

traffic.state.perc.aggr <- data.with.name %>%
  group_by(shortname) %>%
  summarise_at(vars(totfatrte, nghtfatrte,wkndfatrte), funs(mean))

ggplot(traffic.state.perc.aggr, aes(shortname, totfatrte)) +
  geom_bar(aes(fill = factor(shortname)), position = "dodge", stat="identity") + ggtitle("Fatalities by state") +    theme_gray() + 
  scale_fill_hue(c=45,l=80) + theme(plot.title = element_text(size = 8, hjust = 0.5)) + 
  theme(axis.text.x = element_text(angle = 90, size = 6, vjust = 0.5, hjust=1)) +
  geom_line(aes(x = shortname, y = nghtfatrte, color="red"), size = 1.5, group = 1) + 
geom_line(aes(x = shortname, y = wkndfatrte, color="blue"), size = 1.5, group = 1) +
  scale_color_discrete(name = "Rates", labels = c("Night Fatality Rate", "Weekend Fatality Rate"))

```

```{r EDA4}

library(ggnewscale)
library(gridExtra)

#get top and bottom states data, we'll compare these states for traffic law analyses
library(viridis)
top.5.fatalities <- traffic.state.perc.aggr %>%
         filter(rank(desc(totfatrte))<=5) 

bottom.5.fatalities <- traffic.state.perc.aggr %>%
         filter(rank((totfatrte))<=5)

data.top.filtered <- data.with.name %>% filter(shortname %in% c("MS","MT","NM","SC","WY"))
data.bottom.filtered <- data.with.name %>% filter(shortname %in% c("CT","MA","NJ","NY","RI"))

data.merged <- data.with.name %>% filter(shortname %in% c("MS","MT","NM","SC","WY","CT","MA","NJ","NY","RI"))

top.plot.1 <- ggplot(data.top.filtered, aes(x = year, y = totfatrte, group = shortname, color = shortname)) +    scale_color_viridis(discrete = TRUE, option = "D") +  
  theme(plot.title = element_text(size = 10, hjust = 0.5)) + ggtitle("Fatalities by state - Top 5 ") +
  geom_line(aes(x = year,y=totfatrte)) + theme_gray() +
  ggnewscale::new_scale_colour() + 
  geom_point(aes(colour = factor(seatbelt))) + scale_color_discrete()

bottom.plot.1 <- ggplot(data.bottom.filtered, aes(x = year, y = totfatrte, color = factor(shortname) )) + theme_gray() +
  theme(plot.title = element_text(size = 10, hjust = 0.5)) +  scale_color_viridis(discrete = TRUE) +  ggtitle("Fatalities by State - Bottom 5") +   
  geom_line(aes(x = year,y=totfatrte)) + 
  ggnewscale::new_scale_colour() + 
  geom_point(aes(colour = factor(seatbelt))) + scale_color_discrete()

plot.1 <- ggplot(data.merged, aes(x = year, y = totfatrte, group = shortname, color = shortname)) + theme_gray() +
  theme(plot.title = element_text(size = 10, hjust = 0.5)) + scale_color_discrete(name = "State") + ggtitle("Fatalities by Top 5 and Bottom 5 states") +  
  geom_line(aes(x = year,y=totfatrte)) + 
  ggnewscale::new_scale_colour() + 
  geom_point(aes(colour = factor(seatbelt))) + scale_color_viridis(discrete = TRUE, option = "D",name = "Seatbelt") 

plot.1
top.plot.1
bottom.plot.1

grid.arrange(plot.1, top.plot.1, bottom.plot.1, 
             nrow = 3, heights = c(20,20,20), ncol = 1)

```

```{r}
eda.states.chart <- function(fieldname) {
  
top.plot.1 <- ggplot(data.top.filtered, aes(x = year, y = totfatrte, group = shortname, color = shortname)) +  theme_gray() +  theme(plot.title = element_text(size = 10, hjust = 0.5)) + scale_color_viridis(discrete = TRUE, option = "D") +  
   ggtitle("Fatalities by state - Top 5") +
  geom_line(aes(x = year,y=totfatrte)) + theme_gray() +
  ggnewscale::new_scale_colour() + 
  geom_point(aes(colour = factor(get(fieldname)))) + scale_color_discrete(name = fieldname)

bottom.plot.1 <- ggplot(data.bottom.filtered, aes(x = year, y = totfatrte, color = factor(shortname) )) + theme_gray() +
  theme(plot.title = element_text(size = 10, hjust = 0.5)) +  scale_color_viridis(discrete = TRUE) +  ggtitle("Fatalities by state - Bottom 5") +   
  geom_line(aes(x = year,y=totfatrte)) + 
  ggnewscale::new_scale_colour() + 
  geom_point(aes(colour = factor(get(fieldname)))) + scale_color_discrete( name = fieldname)

plot.1 <- ggplot(data.merged, aes(x = year, y = totfatrte, group = shortname, color = shortname)) + theme_gray() +
  theme(plot.title = element_text(size = 10, hjust = 0.5)) + scale_color_discrete(name = "State") + ggtitle("Fatalities by Top 5 and Bottom 5 states") +  
  geom_line(aes(x = year,y=totfatrte)) + 
  ggnewscale::new_scale_colour() + 
  geom_point(aes(colour = factor(get(fieldname)))) + scale_color_viridis(discrete = TRUE, option = "D",name = fieldname) 

plot.1
top.plot.1
bottom.plot.1
# 
# grid.arrange(plot.1, top.plot.1, bottom.plot.1,
#               nrow = 1, heights = c(20), ncol = 3)
  
}

eda.states.chart("bac10")
```
```{r 1.EDA2,warning=FALSE,message=FALSE}

coplot(totfatrte ~ year|state, type="l", data=data) # Lines
coplot(totfatrte ~ year|state, type="b", data=data)

scatterplot(totfatrte ~ year|state, boxplots=FALSE, smooth=TRUE, reg.line=FALSE, data=data)

plotmeans(totfatrte ~ perse, main="Heterogeneity across States", data=data)

plotmeans(totfatrte ~ year, main="Heterogeneity across Years", data=data)
```

```{r usmap,warning=FALSE,message=FALSE}
library(tidyverse)
library(maps)
library(mapproj)
us_states <- map_data("state")

traffic.state.perc.name.aggr <- data.with.name %>%
  group_by(name) %>%
  summarise_at(vars(totfatrte), funs(mean))

traffic.state.perc.name.aggr$name <- tolower(traffic.state.perc.name.aggr$name)

p <- ggplot(data = us_states,
            mapping = aes(x = long, y = lat,
                          group = group, fill = region))

p + geom_polygon() +
    guides(fill = FALSE)


us_states %>% 
  left_join(traffic.state.perc.name.aggr, by=c("region"="name")) %>%
  ggplot(aes(x=long,y=lat,group=group, fill=totfatrte)) +
  geom_polygon(color = "gray90", size = 0.1) +
  #coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  coord_map(projection = "albers", lat0 = 45, lat1 = 55) +
  scale_fill_continuous(type = "viridis")+
  #scale_fill_brewer("Oranges")+
  theme(legend.position="bottom",
        axis.line=element_blank(),
        axis.text=element_blank(),
        axis.ticks=element_blank(),
        axis.title=element_blank(),
        panel.background=element_blank(),
        panel.border=element_blank(),
        panel.grid=element_blank())  + theme_gray() + 
  theme(plot.title = element_text(size = 10, hjust = 0.5)) + ggtitle("Fatality Rates By State")


# library(plotly)
# library(sf)
# g <- ggplot(as.data.frame(traffic.state.perc.aggr)) +
#   geom_sf(aes(fill=totfatrte)) +
#   scale_fill_distiller("Fatality Rate", palette="Spectral") +
#   ggtitle("Fatality Rate By State")
# ggplotly(g)

#str(traffic.state.perc.aggr)
```

2. (15%) How is the our dependent variable of interest *totfatrte* defined? What is the average of this variable in each of the years in the time period covered in this dataset? Estimate a linear regression model of *totfatrte* on a set of dummy variables for the years 1981 through 2004. What does this model explain? Describe what you find in this model. Did driving become safer over this period? Please provide a detailed explanation.

3. (15%) Expand your model in *Exercise 2* by adding variables *bac08, bac10, perse, sbprim, sbsecon, sl70plus, gdl, perc14_24, unem, vehicmilespc*, and perhaps *transformations of some or all of these variables*. Please explain carefully your rationale, which should be based on your EDA, behind any transformation you made. If no transformation is made, explain why transformation is not needed. How are the variables *bac8* and *bac10* defined? Interpret the coefficients on *bac8* and *bac10*. Do *per se laws* have a negative effect on the fatality rate? What about having a primary seat belt law? (Note that if a law was enacted sometime within a year the fraction of the year is recorded in place of the zero-one indicator.)

4. (15%) Reestimate the model from *Exercise 3* using a fixed effects (at the state level) model. How do the coefficients on *bac08, bac10, perse, and sbprim* compare with the pooled OLS estimates? Which set of estimates do you think is more reliable? What assumptions are needed in each of these models?  Are these assumptions reasonable in the current context?

5. (10%) Would you perfer to use a random effects model instead of the fixed effects model you built in *Exercise 4*? Please explain.

6. (10%) Suppose that *vehicmilespc*, the number of miles driven per capita, increases by $1,000$. Using the FE estimates, what is the estimated effect on *totfatrte*? Please interpret the estimate.

7. (5%) If there is serial correlation or heteroskedasticity in the idiosyncratic errors of the model, what would be the consequences on the estimators and their standard errors?


#TODO in EDA mention about omitted variables - for example advances in medical field, ER procedures. Also compliance to traffic laws - seatbelt, drunk driving. Also Road condition. For example Wyoming and SC have large rural roads. 

#TODO in conclusion 
 The researchers at the Ann Arbor based institute found that while road fatalities were lower than other causes of death, they were still considerable. Michael Sivak and Brandon Schoettle, authors of the report "Mortality from road crashes in the individual U.S. states: A comparison with leading causes of death," found that nationwide, death from road crashes per 100,000 population was 10.9, compared to 34.4 for Alzheimer's, 43.7 for stroke, 48.2 for lung disease, 185.4 from cancer and 197.2 from heart disease. They used 2015 data from the National Highway Traffic Safety Administration and the Centers for Disease Control and Prevention
 Traffic fatalities from crashes in the United States are lower than other leading causes of death, but still substantial. The Northeast may be the safest region in the country on the roads; Wyoming is the most dangerous state.










